# Chapter 6: Analysis of longitudinal data
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Chapter 8 of MABS using the RATS 

Read in the data (from the RDS object that retains the factor type definition):

```{r}
library(data.table)
library(ggplot2)
library(dplyr)
library(magrittr)
RATSL<-readRDS("data/RATSL.rds")
```

### Glimpse the data

```{r}
glimpse(RATSL)
head(RATSL)
tail(RATSL)
```

So we have rats with multiple observations and each one belongs to a one of three treatment groups. Next let's plot weight of each rat in each groups separately as a time series:

### Plot groups separately

```{r}
p1 <- ggplot(RATSL, aes(x = Time, y = Weight, linetype = ID))
p2 <- p1 + geom_line() + scale_linetype_manual(values = rep(1:10, times=4))
p3 <- p2 + facet_grid(. ~ Group, labeller = label_both)
p4 <- p3 + theme_bw() + theme(legend.position = "none")
p5 <- p4 + theme(panel.grid.minor.y = element_blank())
p6 <- p5 + scale_y_continuous(limits = c(min(RATSL$Weight), max(RATSL$Weight)))
p6
```

We can already see that there is a increasing trend in all groups and already a significant baseline differencies among groups.

### Standardise the scores and replot

```{r}
RATSL <- RATSL %>%
   group_by(Time) %>%
   mutate( stdweight = (Weight - mean(Weight))/sd(Weight) ) %>%
   ungroup()
glimpse(RATSL)
```


```{r}
p1 <- ggplot(RATSL, aes(x = Time, y = stdweight, linetype = ID))
p2 <- p1 + geom_line() + scale_linetype_manual(values = rep(1:10, times=4))
p3 <- p2 + facet_grid(. ~ Group, labeller = label_both)
p4 <- p3 + theme_bw() + theme(legend.position = "none")
p5 <- p4 + theme(panel.grid.minor.y = element_blank())
p6 <- p5 + scale_y_continuous(name = "standardized weight")
p6
```

After standardizing, we can see some differences in trends: Group 1 seems to be quite steady, group 2 mostly increasing weigth (with one exception) and group 3 generally has a decreasing trend.

### Number of weeks, baseline (week 0) included:

```{r}
n <- RATSL$Time %>% unique() %>% length()
# Make a summary data:
RATSS <- RATSL %>%
   group_by(Group, Time) %>%
   summarise( mean=mean(Weight), se=sd(Weight)/sqrt(n) ) %>%
   ungroup()
glimpse(RATSS)
```

```{r}
p1 <- ggplot(RATSS, aes(x = Time, y = mean, linetype = Group, shape = Group))
p2 <- p1 + geom_line() + scale_linetype_manual(values = c(1,2,3))
p3 <- p2 + geom_point(size=3) + scale_shape_manual(values = c(1,2,3))
p4 <- p3 + geom_errorbar(aes(ymin=mean-se, ymax=mean+se, linetype="1"), width=0.3)
p5 <- p4 + theme_bw() + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())
p6 <- p5 + theme(legend.position = c(0.8,0.8))
p7 <- p6 + scale_y_continuous(name = "mean(Weight) +/- se(Weight)")
p7
```

Groups have clear baseline differences. We can also see that variation inside groups is different (group 1 have very similar weights, group 2 has largest in-group differences). Generally all groups show increasing trend, group 2 maybe highest.

### Boxplot to spot incosistent distributions

```{r}
p1 <- ggplot(RATSL, aes(x = factor(Time), y = Weight, fill = Group))
p2 <- p1 + geom_boxplot(position = position_dodge(width = 0.9))
p3 <- p2 + theme_bw() + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())
p4 <- p3 + theme(legend.position = c(0.8,0.8))
p5 <- p4 + scale_x_discrete(name = "week")
# Black & White version:
#p6 <- p5 + scale_fill_grey(start = 0.5, end = 1)
p5
```

We can see that group 2 actually seems to have one outlier that causes mean to go higher than maybe reasonable. Also group 1 has one outlier that drags mean downwards. If we compare trends of group medians (that is more robust to utliers), they show quite similar increasing slopes.

### Group differences and boxplots

```{r}
# Make a summary data of the post treatment weeks (1-8)
RATSG <- RATSL %>%
   filter(Time > 0) %>%
   group_by(Group, ID) %>%
   summarise( mean=mean(Weight) ) %>%
   ungroup()
glimpse(RATSG)
```

```{r}
p1 <- ggplot(RATSG, aes(x = Group, y = mean))
p2 <- p1 + geom_boxplot()
p3 <- p2 + theme_bw() + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())
p4 <- p3 + stat_summary(fun.y = "mean", geom = "point", shape=23, size=4, fill = "white")
p5 <- p4 + scale_y_continuous(name = "mean(Weight)")
p5
```

Same outliers that we identified earlier show up here also. We also now find one outlier in group 3, that earlier could have been considered as an borderline case (appearing only on three weeks)

### Remove outliers

```{r}
RATSG1 <- RATSG %>%
   filter( (Group == 1 & mean > 240) |
             (Group == 2 & mean < 500)  |
             (Group == 3 & mean > 500))
glimpse(RATSG1)
```

```{r}
p1 <- ggplot(RATSG1, aes(x = Group, y = mean))
p2 <- p1 + geom_boxplot()
p3 <- p2 + theme_bw() + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())
p4 <- p3 + stat_summary(fun.y = "mean", geom = "point", shape=23, size=4, fill = "white")
p5 <- p4 + scale_y_continuous(name = "mean(Weight)")
p5
```

Now that outliers are removed, variances inside groups look much better.

### ANOVA without outliers 

Without the outliers, let's apply one-way analysis of variance (ANOVA is a generalization of t-test. T-test works only with only in case of two groups)

```{r}
summary(aov(mean ~ Group, data = RATSG1))
```

We see highly statistically highly significant difference among groups (p=2.7e-14) which was anticipated from the earlier plots already.

### Include baseline

Next, add the baseline from the original data as a new variable to the summary data, and then remove outliers again.

```{r warning=F}
RATS<-fread("https://raw.githubusercontent.com/KimmoVehkalahti/MABS/master/Examples/data/rats.txt")[,-1,with=F] # RATS has row numbers, remove them
baseline <- RATS$WD1
RATSG2 <- RATSG %>% mutate(baseline) %>%
   filter( (Group == 1 & mean > 240) |
             (Group == 2 & mean < 500)  |
             (Group == 3 & mean > 500))

```

### Fit the ANCOVA model and see the results

```{r}
fit <- lm(mean ~ baseline + Group, data = RATSG2)
summary(fit)
```

Including baseline as covariate, we can see statistical significance among all three groups, but it's not as huge as without baseline.

## Chapter 9 of MABS using the BPRS

Read in the data (from the RDS object that retains the factor type definition):

```{r}
BPRSL<-readRDS("data/BPRSL.rds")
glimpse(BPRSL)
```

### Plot

```{r}
p1 <- ggplot(BPRSL, aes(x = Week, y = bprs, group = subject))
p2 <- p1 + geom_text(aes(label = treatment))
p3 <- p2 + scale_x_continuous(name = "Time (week)", breaks = seq(0, 60, 10))
p4 <- p3 + scale_y_continuous(name = "BPRS")
p5 <- p4 + theme_bw()
p6 <- p5 + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())
p6
```

Hard to make any conclusions, but general trend seems to be declining and slightly increasing at the end.

### Regression

```{r}
BPRS_reg <- lm(bprs ~ Week + treatment, data = BPRSL)
summary(BPRS_reg)
```

There is a statistically significant trend in time, but treatment doesn't seem to associate.

### Plot

```{r}
p1 <- ggplot(BPRSL[order(Week)], aes(x = Week, y = bprs, group = interaction(subject, treatment)))
p2 <- p1 + geom_line() # + geom_line(aes(linetype = treatment)) # treatment given in middle
p3 <- p2 + scale_x_continuous(name = "Time (week)", breaks = seq(0, 60, 10))
p4 <- p3 + scale_y_continuous(name = "BPRS")
p5 <- p4 + theme_bw() + theme(legend.position = "top")
p6 <- p5 + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())
p6
```

Line plot confirms this association with time, as well as pair plots below.

### Pairs plot

```{r}
BPRS<-fread("https://raw.githubusercontent.com/KimmoVehkalahti/MABS/master/Examples/data/BPRS.txt")
pairs(BPRS[, 3:11], cex = 0.7)
```

### LMM, random intercept

Let's analyze data now with linear mixed model and three different model definitions: random intercept, random slope and intercept, and random slope and random intercept including their interaction term.

```{r}
library(lme4)
BPRS_ref <- lmer(bprs ~ Week + treatment + (1 | subject), data = BPRSL, REML = FALSE)
summary(BPRS_ref)
```

### LMM, random slope and intercept
```{r}
BPRS_ref1 <- lmer(bprs ~ Week + treatment + (Week | subject), data = BPRSL, REML = FALSE)
summary(BPRS_ref1)
anova(BPRS_ref, BPRS_ref1)
```

Adding slope terms seems to make the model somewhat better (p=0.03).

### LMM, random slope and intercept, intercation term
```{r}
BPRS_ref2 <- lmer(bprs ~ Week * treatment + (Week | subject), data = BPRSL, REML = FALSE)
summary(BPRS_ref2)
anova(BPRS_ref1, BPRS_ref2)
```

Adding interaction doesn't seem to be so important (p=0.07) but probably doesn't hurt either.

### Plot fitted values from models
```{r}
Fitted <- fitted(BPRS_ref2)
BPRSL <- BPRSL %>% mutate(Fitted)
p1 <- ggplot(BPRSL, aes(x = Week, y = bprs, group = interaction(subject, treatment)))
p2 <- p1 + geom_line() #+ geom_line(aes(linetype = Group))
p3 <- p2 + scale_x_continuous(name = "Time (week)", breaks = seq(0, 60, 20))
p4 <- p3 + scale_y_continuous(name = "BPRS")
p5 <- p4 + theme_bw() + theme(legend.position = "right") # "none" in the book
p6 <- p5 + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())
p7 <- p6 + ggtitle("Observed")
graph1 <- p7

p1 <- ggplot(BPRSL, aes(x = Week, y = Fitted, group = interaction(subject, treatment)))
p2 <- p1 + geom_line() #+ geom_line(aes(linetype = Group))
p3 <- p2 + scale_x_continuous(name = "Time (week)", breaks = seq(0, 60, 20))
p4 <- p3 + scale_y_continuous(name = "BPRS")
p5 <- p4 + theme_bw() + theme(legend.position = "right")
p6 <- p5 + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())
p7 <- p6 + ggtitle("Fitted")
graph2 <- p7

graph1; graph2
```

Fitted values seem to estimate general declining linear trend for all individuals. However, as we noted in the earlier first plot, the relationship is probably actually non-linear, meaning it increases at the end. Thus, depending on application, we might want to include some second order polynomial term to account for this non-linearity:

```{r}
BPRS_ref3 <- lmer(bprs ~ poly(Week,2) + treatment + (Week | subject), data = BPRSL, REML = FALSE)
summary(BPRS_ref3)
anova(BPRS_ref2, BPRS_ref3)
```

It seems to perform much better. Let' plot these fitted values with second order polynomial term:
```{r}
Fitted <- fitted(BPRS_ref3)
BPRSL2 <- BPRSL %>% mutate(Fitted)
p1 <- ggplot(BPRSL2, aes(x = Week, y = bprs, group = interaction(subject, treatment)))
p2 <- p1 + geom_line() #+ geom_line(aes(linetype = Group))
p3 <- p2 + scale_x_continuous(name = "Time (week)", breaks = seq(0, 60, 20))
p4 <- p3 + scale_y_continuous(name = "BPRS")
p5 <- p4 + theme_bw() + theme(legend.position = "right") # "none" in the book
p6 <- p5 + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())
p7 <- p6 + ggtitle("Observed")
graph1 <- p7

p1 <- ggplot(BPRSL2, aes(x = Week, y = Fitted, group = interaction(subject, treatment)))
p2 <- p1 + geom_line() #+ geom_line(aes(linetype = Group))
p3 <- p2 + scale_x_continuous(name = "Time (week)", breaks = seq(0, 60, 20))
p4 <- p3 + scale_y_continuous(name = "BPRS")
p5 <- p4 + theme_bw() + theme(legend.position = "right")
p6 <- p5 + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())
p7 <- p6 + ggtitle("Fitted")
graph2 <- p7

graph1; graph2
```

=> Profit!
